Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-09-23 inst/doc/nonparametric-bayes.Rmd
tackling intro yet again

<!--
Very briefly set the stage: (see related papers)
- Decision-theory / optimal control framework
- Fisheries context 
-->

<!-- FIXME -->

@@

Uncertainty enters at many levels <!-- enters what -->
management

- Why fisheries

<!-- Other reasons for failure see Sethi-->

@Fischer 
@Sethi2005, @Peretti, @Munch,

(Classical division)

- Dynamic uncertainty (stochasticity)

From [Clark (1976)]() to [Reed (1979)]()

Other sources of uncertainty [Roughgarden & Smith 1996](), [Sethi _et al._ (2005)]()

- Parametric uncertainty <!-- ugh, be more consise and less annoying-->

As the parameter values for these models must be estimated from limited data,  
there will always be some uncertainty associated with these values.  This uncertainty
further compounds the intrinsic variability introduced by demographic or environmental 
noise.  The degree of uncertainty in the parameter values can be inferred from the data
and reflected in the estimates of the transition probabilities.

[@Mangel1985; @Schapaugh2013]

[Ludwig & Walters 1982](http://doi.org/10.1016/0304-3800(82)90023-0 "Optimal harvesting with imprecise parameter estimates")

- Structural uncertainty <!-- ugh, be more consise and less annoying-->

@@

[@Cressie2009](http://dx.doi.org/10.1890/07-0744.1 "Accounting for uncertainty in ecological analysis: the strengths and limitations of hierarchical statistical modeling.").

@@

The three sources of uncertainty: stochasticity (in process, measurement, and implementation; though we focus on process-noise only for simplicity), parametric uncertainty, and structural uncertainty, are familiar challenges.

Of all sources of uncertainty, this last has received the least attention.[^1]

Often the set of feasible states is larger than the range of the observed states.  Though most systems subject to some degree of stochasticity can enter a state outside previously observed values by chance alone (such as a particularly low or high nutrient year), this concern is most keenly felt in the decision-theoretic context. If we are observing the fluctuations of a given fish stock over many years under a fixed harvesting pressure, we might develop and test a model that could reasonably predict the frequency of a deviation of a given size, even when such a deviation has not been previously observed.

(The same analogy might be made to predicting hurricanes or other extreme weather events from historical data, which may be more accurately predicted in a constant climate than in one being perturbed).

<!-- What do we call this?  Extrapolation uncertainty?  Pathological Uncertainty? -->

@@

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation. 
Here we illustrate how a stochastic dynamic programming (SDP) algorithm can 
be driven by the predictions from a Gaussian Process (GP).
This provides two distinct advantages compared with contemporary approaches.
First, using a GP sidesteps the need for an accurate model-based description
of the system dynamics (or a model choice or model-averaging approach which
introduces yet more difficulty).  
Second, unlike parametric models which can only reflect 
uncertainty in the corresponding confidence intervals or posterior distributions
of their parameters, the GP approach provides a state-space dependent representation
of uncertainty.  This permits a much greater uncertainty far from the observed data than near the observed 
data.  These features allow the GP-SDP approach to find robust management solutions 
in face of limited data and without knowledge of the correct model structure.

@@

<!--move to  much later --> 
- Note on terminology: "Non-parametric"

The term non-parametric to describe Gaussian processes (GP) is unfortunate,
given that the GP models still involve the estimation of parameters. Some 
literature thus uses the term semi-parametric, which merely adds ambiguity 
to the confusion.  Statisticians give two distinct meanings to the term: (a)
referring to a method that does not assume a distribution (e.g. non-parametric bootstrap, order statistics),
and (b) one that does not assume a fixed structure but rather depends on the 
size of the data (e.g. non-parametric regression, non-parametric hierarchical models
such as the Dirichlet process).  Our use is in the latter sense.

This non-parametric property -- having a structure that is not fixed but 
rather dependent on the data is precisely the property that makes this approach
attractive in face of the limited data sampling challenges discussed above.  
Having fit a parametric model to some data, the model is completely described
by the values (or posterior distributions) of it's parameters.  The non-parametric 
model is not captured by its parameter values or distributions alone -- the 
data points become part of the model.

@@

- Definition
- Previous application
- Why it is particularly suited to these two problems
- (Why this is a novel application thereof)

@@

Given parameters for the function $f$ and probability distribution $Z$,
along with a given economic model determining the  price/profit $\Pi(X_t,
h_t)$ realized in a given year given a choice of harvest $h_t$ and
observed stock $X_t$.  This problem can be solved exactly for discretized
values of stock $X$ and policy $h$ using stochastic dynamic programming
(SDP) [@Mangel1985]. Problems of this sort underpin much marine fisheries
management today.

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Uncertainty in the parameter estimates can 
be estimated and integrated over to determine the optimal policy under
under uncertainty [@Mangel1985; @Schapaugh2013]. Uncertainty in the model
structure itself can only be addressed in this approach by hypothesizing 
alternative model structures, and then performing some model choice or
model averaging  [@Williams2001; @Athanassoglou2012].

@@

The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005] and introduced
in the ecological modeling and fisheries management by @Munch2005.
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.

>>>>

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.

@@

<!-- Uncertainty outside the data without the correct model has not been handled. -->

Uncertainty enters the decision-making process at many levels: intrinsic 
stochasticity in biological processes, measurements, and implementation of 
policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1997; @Sethi2005], parameteric 
uncertainty [_e.g._ @Ludwig1982; @Hilborn1997; @McAllister1998; @Schapaugh2013],
and model or structural uncertainty [_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].
Of these, structural uncertainty incorporates the least a priori knowledge or 
assumptions and is generally the hardest to quantify. Typical approaches assume 
a weak notion of uncertainty where a correct or reasonable approximation of the dynamics
must be identified from among a handful of alternative models.  Here we consider an approach
that addresses uncertainty at each of these levels without assuming the dynamics follow
a particular (i.e. parametric) structure.

<!-- What do we call this?  Extrapolation uncertainty?  Pathological Uncertainty? -->
An additional source of uncertainty that has recieved less attention[^1] arises <!-- remove 'typical'? -->
when applying a dynamical model outside the range of data on which it has been
estimated.  This extrapolation uncertainty is felt most keenly in decision-theoretic
applications, as (a) exploring the potential action space typically involves considering
actions that may move the system outside the range of observed behavior, and (b) decision-theoretic
alogrithms rely not only on reasonable estimates of the expected outcomes, but depend on
the weights given to all possible outcomes [_e.g._ @Weitzman2013].

_Map_

Here we illustrate how a stochastic dynamic programming (SDP) algorithm [@Mangel1988; @Possingham1997, Marescot2013] can 
be driven by the predictions from a Bayesian non-parametric (BNP) approach [@Munch2005a].  
This provides two distinct advantages compared with contemporary approaches.
First, using a BNP sidesteps the need for an accurate model-based description
of the system dynamics.  
Second, the BNP can better reflect uncertainty that arises when extrapolating a model 
outside of the data on which it was fit.  
We illustrate that when the correct model is not known, this latter
feature is crucial to providing a robust decision-theoretic approach in
face of substantial structural uncertainty.

Often the set of feasible states is larger than the range of the observed
states.  Though most systems subject to some degree of stochasticity
can enter a state outside previously observed values by chance alone
(such as a particularly low or high nutrient year), this concern is
most keenly felt in the decision-theoretic context. If we are observing
the fluctuations of a given fish stock over many years under a fixed
harvesting pressure, we might develop and test a model that could
reasonably predict the frequency of a deviation of a given size, even
when such a deviation has not been previously observed.
(The same analogy might be made to predicting hurricanes or other
extreme weather events from historical data, which may be more accurately
predicted in a constant climate than in one being perturbed).

<!-- More Consise versions above

- Parametric uncertainty

As the parameter values for these models must be estimated from limited data,  
there will always be some uncertainty associated with these values.  This uncertainty
further compounds the intrinsic variability introduced by demographic or environmental 
noise.  The degree of uncertainty in the parameter values can be inferred from the data
and reflected in the estimates of the transition probabilities [@Walters1982; @Mangel1988; @Mangel1997; @Schapaugh2013].

- Structural uncertainty

@@

@@

@@

<!-- necessary? -->
This paper represents the first time the SDP decision-making framework has been used
without an a priori model of the underlying dynamics through the use of the BNP approach.

In contrast to parametric models which can only reflect uncertainty 
in parameter estimates, the BNP approach provides a more state-space dependent representation
of uncertainty.

This permits a much greater uncertainty far from the observed data than near the observed 
data.  These features allow the GP-SDP approach to find robust management solutions 
in face of limited data and without knowledge of the correct model structure.

@@

- Note on: Why fisheries

<!--move to  much later -->

@@

- Background on non-parametric modeling.

Addressing the difficulty posed by extrapolation without knowing the
true model requires a nonparametric approach to model fitting: one
that does not assume a fixed structure but rather depends on the size
of the data (e.g. non-parametric regression or a Dirichlet process).
This established terminology is nevertheless unfortunate, as (a)
this approach still involves the estimation of parameters, and (b),
Statisticians use non-parametric to mean both this property (structure
is not fixed by the parameters) and an entirely different (and probably
more familiar) case in which the model does not assume any distribution
(e.g. non-parametric bootstrap, order statistics).  Some literature
thus uses the term semi-parametric, which merely adds ambiguity to
the confusion.

This non-parametric property -- having a structure explicitly dependent on
the data -- is precisely the property that makes this approach attractive
in face of the limited data sampling challenges discussed above.
Having fit a parametric model to some data, the model is completely
described by the values (or posterior distributions) of it's parameters.
The non-parametric model is not captured by its parameter values or
distributions alone. Either the model scales with the complexity of the data
on which it is estimated (e.g. nonparametric heirarchical approaches such
as the Dirchlet process) or the data points become themselves part of the model
specification, as in the nonparametric regression used here.  we shall see here.

- Definition

- Previous application

The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005], and was first introduced
in the context ecological modeling and fisheries management in @Munch2005.
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.

- Why it is particularly suited to these two problems
- (Why this is a novel application thereof)

@@

Given parameters for the function $f$ and probability distribution $Z$,
along with a given economic model determining the  price/profit $\Pi(X_t,
h_t)$ realized in a given year given a choice of harvest $h_t$ and
observed stock $X_t$.  This problem can be solved exactly for discretized
values of stock $X$ and policy $h$ using stochastic dynamic programming
(SDP) [@Mangel1988]. Problems of this sort underpin much marine fisheries
management today.

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Uncertainty in the parameter estimates can 
be estimated and integrated over to determine the optimal policy under
under uncertainty [@Mangel1988; @Schapaugh2013]. Uncertainty in the model
structure itself can only be addressed in this approach by hypothesizing 
alternative model structures, and then performing some model choice or
model averaging  [@Williams2001; @Athanassoglou2012].

@@
